---
title: "General introduction"
author: "Carl Brunius, Vilhelm Suksi"
date: "`r Sys.Date()`"
output:
  BiocStyle::html_document:
    toc: true
    toc_depth: 2
    number_sections: true
vignette: >
  %\VignetteIndexEntry{General introduction}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
bibliography: references.bib
biblio-style: apalike
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "##")
```

# Motivation

From the perspective of metabolites as the continuation of the central dogma of biology, metabolomics provides the closest link to many phenotypes of interest. This makes metabolomics research promising in teasing apart the complexities of living systems. However, due to experimental reasons, the data includes non-biological variation which limits quality and reproducibility, especially if the data is obtained from several batches. 

The batchCorr package reduces unwanted variation by way of between-batch alignment, within-batch drift correction and between-batch normalization using batch-specific quality control (QC) samples and long-term reference QC samples. Please see the associated article [@brunius2016large] for more thorough descriptions of algorithms.

# Installation

To install ```batchCorr```, install BiocManager first, if it is not installed. Afterwards use the install function from BiocManager.

```{r, eval = FALSE}
if (!requireNamespace("BiocManager", quietly = TRUE))
    install.packages("BiocManager")
BiocManager::install("batchCorr")
```

# Data

The example data allows for demonstration of all core ```batchCorr``` functionality: between-batch alignment, within-batch drift correction and between-batch normalization. The example data from a single analytical mode consists of two objects: B_PT (abundances, matrix) and B_meta (sample and feature metadata, data.frame). 

```{r}
library(batchCorr)
data("ThreeBatchData")
```

# How it works

```batchCorr``` is designed to work with basic data classes. Adundances are included in a matrix, while sample and feature metadata is included in a data.frame. ```batchCorr``` works best as per the chronology presented below. Other functions include:

- ```peakInfo``` to extract m/z and rt from peak table
- ```mergeBatches``` for merging batches after drift correction

Important analytical background includes batch-specific QC samples and long-term reference QC samples, which are regularly interspersed in the injection sequence. Batch-specific QC samples are typically pooled aliquots of study samples, and are used for within-batch drift correction. Long-term reference QC samples are not of the same biological origin as the batch-specific QC samples, and are therefore not directly representative of the sample population. Long-term reference QC samples are used for between-batch alignment, within-batch drift correction and between-batch normalization.

## Between-batch alignment

Shifts in retention time (RT) and mass-to-charge ratio (m/z) across batches results in some metabolites being redundantly represented in the dataset. To rectify this, between-batch alignment of features is performed using ```alignBatches```, which encompasses the following steps:

1. Aggregation of feature presence/missingness on batch level (```batchFlag```)
- batch-wise flagging of low-quality features with proportion of ```NA```s to all samples > 80% based on long-term reference QC samples
- 0 < total batch presence of candidates features < number of batches to be an alignment candidate
2. Identification of features with missingness within "the box", i.e. sufficiently similar in RT and m/z (```alignIndex```) 
- potential alignment candidates have similar RT and m/z across batches
- orthogonal batch presence: two or more alignment candidates cannot be present in the same batch
- if there are multiple combinations of candidates across batches, the features are recursively subclustered before clustering across batches
- saves plots for inspecting results of subclustering

3. Plotting clusters of alignment candidates (```plotAlign```)

4. Aligning of feature clusters resulting in new peak table  (```batchAlign```)

```{r}
## Perform batch alignment
# Extract peakinfo (i.e. m/z and rt of features), 
# These column names have 2 leading characters describing LC-MS mode 
# -> start at 3
peakIn <- peakInfo(PT = PTnofill, sep = '@', start = 3) 
# Perform multi-batch alignment
alignBat <- alignBatches(peakInfo = peakIn, PeakTabNoFill = PTnofill,
                         PeakTabFilled = PTfill, batches = meta$batch, 
                         sampleGroups = meta$grp, selectGroup = 'QC')
# Extract new peak table
PT=alignBat$PTalign
```

## Within-batch drift correction
Drift in abundance within a batch gives rise to unwanted variation which can be modelled in terms of injection order. Many methods fail to take into account different drift patterns in features or are prone to overfitting. Herein, within-batch drift correction is performed using the wrapper ```correctDrift```, which involves:

1. Clustering of features (```clust```)
- scaling by standard deviation
- clustering serves to identify features with similar drift patterns, which are corrected in aggregate. As such, different drift patterns are accounted for while mitigating overfitting to unwanted variation in a single feature. 
- different cluster model types (```groupNames```) and numbers of clusters (```G```) can be tested 

2. Calculation of correction factor (```dritCalc```)
- ```driftCalc```accepts arguments specifying the smoothing function (LOESS or cubic spline)

3. Correction of the abundances using correction factor (```driftCorr```)
- corrects to reference level at the first injection after scaling
- plots the feature clusters against injection order before and after drift correction
- corrected values were retained only if the root mean square deviation of long-term reference QC samples was reduced after drift correction for the cluster at large

4. Removal of individual features (```cleanVar```)
- features were removed batch-wise if the CV of within-batch QC samples after drift correction > 30%


```{r, results = 'hide', output = FALSE}
# Batch B
batchB <- getBatch(peakTable = PT, meta = meta, 
                   batch = meta$batch, select = 'B')
BCorr <- correctDrift(peakTable = batchB$peakTable, 
                      injections = batchB$meta$inj, 
                      sampleGroups = batchB$meta$grp, QCID = 'QC', 
                      G = seq(5,35,by=3), modelNames = c('VVE', 'VEE'))
# Batch F
batchF <- getBatch(peakTable = PT, meta = meta, 
                   batch = meta$batch, select = 'F')
FCorr <- correctDrift(peakTable = batchF$peakTable, 
                      injections = batchF$meta$inj, 
                      sampleGroups = batchF$meta$grp, 
                      QCID = 'QC', G = seq(5,35,by=3), 
                      modelNames = c('VVE', 'VEE'))
# Batch H
batchH <- getBatch(peakTable = PT, meta = meta, 
                   batch = meta$batch, select = 'H')
HCorr <- correctDrift(peakTable = batchH$peakTable, 
                      injections = batchH$meta$inj, 
                      sampleGroups = batchH$meta$grp, 
                      QCID = 'QC', G = seq(5,35,by=3),
                      modelNames = c('VVE', 'VEE'))
```

## Between-batch normalization

```normalizeBatches``` performs between-batch normalisation either based on long-term reference QC samples or median batch intensity depending on the following dual criterion:

1. long-term reference QC sample CV < 30%
2. fold-change < 5 for the ratio of the average feature intensity of a specific feature between batches to the ratio of the all-feature average intensity between batches

If the long-term QC samples are not considered reliable according to the above dual criterion for a specific feature, batches were normalised by sample population median, where a sample population can be specified explicitly to the ```population``` argument. Features not present in all batches are also excluded from the dataset.

```{r, warning = FALSE}
mergedData <- mergeBatches(list(BCorr,FCorr,HCorr))
normData <- normalizeBatches(peakTableCorr = mergedData$peakTableCorr, 
                             batches = meta$batch, sampleGroup = meta$grp,
                             refGroup = 'Ref', population = 'all')
PTnorm <- normData$peakTable
```

# Authors & Acknowledgements

The first version of ```batchCorr``` was written by Carl Brunius, Lin Shi and Rikard Landberg. The package is inspired by...
```batchCorr``` was developed for Bioconductor by Carl Brunius, Vilhelm Suksi and ...
If you find any bugs or other things to fix, please submit an issue on GitLab! All contributions to the package are always welcome!

# Session information

```{r, echo = FALSE}
sessionInfo()
```

# References